---
title: "Example forecast hub data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example forecast hub data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 5,
  fig.width = 8,
  fig.align = "center"
)
```

```{r setup}
library(hubExamples)
library(hubVis)
library(dplyr)
library(ggplot2)
```

# Overview

The `hubExamples` package provides three data sets that contain example model output and
target data for an example forecast hub: `forecast_outputs`, `forecast_target_ts`, and
`forecast_target_observations`. These forecasts and target data are a subset of the model outputs and target data that are provided in the [example-complex-forecast-hub](https://github.com/Infectious-Disease-Modeling-Hubs/example-complex-forecast-hub). These data were in turn derived from forecast submissions and target data for the [FluSight Forecast Hub](https://github.com/cdcepi/Flusight-forecast-data) for the 2022/23 season.

We begin with a high level overview of these data objects and then we describe the different forecast targets in more detail.

## Example forecast output data

The example forecasts provided in `forecast_outputs` are derived from forecasts that were submitted to the FluSight hub from three models: `Flusight-baseline`, `MOBS-GLEAM_FLUH`, and `PSI-DICE`. The original forecasts submitted to the hub were in quantile format, but we have modified those submissions to provide examples of additional model output types and targets. The predictions for these other output types should be viewed only as illustrations of the data formats, not as real examples of forecasts. We will describe the methods used for creating other forecast output types below.

The snippet below shows the format of the `forecast_outputs`.

```{r}
head(forecast_outputs)
str(forecast_outputs)
```

This is a data frame with four groups of columns (see the [hubverse documentation](https://hubverse.io/en/latest/user-guide/model-output.html) for more information about these data formats):

1. The `model_id` identifies the model that produced the predictions.
2. Together, the `location`, `reference_date`, `horizon`, `target_end_date`, and `target` columns serve to identify a prediction task:
    - The `location` column contains a FIPS code specifying the location being predicted.
    - The `reference_date` is a date in ISO format that gives the Saturday ending the week the predictions were generated.
    - The `horizon` gives the difference between the `reference_date` and the target date of the forecasts (`target_end_date`, see next item) in units of weeks. Informally, this describes "how far ahead" the predictions are targeting.
    - The `target_end_date` is a date in ISO format that gives the Saturday ending the week being predicted. For example, if the `target_end_date` is `"2022-12-17"`, predictions are for a quantity relating to influenza activity in the week from Sunday, December 11, 2022 through Saturday, December 17, 2022.
    - The `target` describes the target quantity for the prediction. In the above example, the `target` of `"wk flu hosp rate"` is the weekly rate of hospital admissions per 100,000 population. The targets included in this example will be described in other sections below.
3. The `output_type` and `output_type_id` columns provide metadata about the model predictions.
    - The `output_type` specifies the representation of the predictive distribution.
    - The `output_type_id` gives additional identifying information about the predictions; the information in this column is specific to the `output_type`.
4. The `value` column contains the value of the model's prediction.

The original hub submissions contained predictions for many locations and dates, and quantile forecasts were provided at 23 different quantile levels ranging from 0.01 to 0.99.  To make the example data more manageable, the `forecast_outputs` object contains a subset of these outputs for two locations (Massachusetts, FIPS code `"25"`, and Texas, FIPS code `"48"`) and two reference dates (2022-11-19 and 2022-12-17).  Additionally, for the quantile forecasts we have subset to seven quantile levels: 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, and 0.95.

## Example forecast target data

All predictions are for targets that are based on influenza hospital admissions as reported in the US National Healthcare Safety Network (NHSN). The `forecast_target_ts` object contains the observed values of these hospital admissions in a time series format:

```{r}
head(forecast_target_ts)
tail(forecast_target_ts)
```

The `forecast_target_observations` object contains the observed values for the prediction targets:

```{r}
head(forecast_target_observations)
```

This data frame has a subset of the columns in the `forecast_outputs` that is sufficient to identify the observed value corresponding to each prediction, including the `location`, `target_end_date`, `target`, `output_type`, and `output_type_id`, along with the observed target values, recorded in the `observation` column. Note that the `reference_date` and `horizon` columns are not needed in this data frame, since the `target_end_date` is sufficient to align observations with predictions.

# Further detail on the forecast targets

The example forecast data contains the following combinations of `target` and `output_type`:

```{r}
forecast_outputs |> distinct(target, output_type)
```

We will describe each of these targets in the following sections.

## The `wk inc flu hosp` target

The `wk inc flu hosp` target represents weekly new hospital admissions with a confirmed influenza diagnosis.  We have predictions of this target with three output types: `quantile`, `mean`, and `median`.  The following plot shows the quantile and median predictions along with the observed hospital admission counts for Massachusetts and Texas.  Note that the quantile predictions were contributed directly by modelers to the FluSight hub, and median predictions correspond exactly to the quantile predictions at probability level 0.5.  We have obtained mean predictions from the quantile forecasts using the [distfromq package](https://github.com/reichlab/distfromq) by estimating the full quantile function from the submitted quantile predictions, drawing a sample using the probability integral transform method, and computing the mean of those samples.

```{r}
plot_step_ahead_model_output(
  model_output_data = forecast_outputs |>
    filter(output_type %in% c("quantile", "median")),
  target_data = forecast_target_ts |>
    filter(location %in% c("25", "48"),
           date >= "2022-10-01", date <= "2023-04-01"),
  use_median_as_point = TRUE,
  x_col_name = "target_end_date",
  intervals = c(0.5, 0.8, 0.9),
  facet = "location",
  group = "reference_date",
  interactive = FALSE
)
```

For purposes of evaluating predictions, it can be helpful to join the observed target values, contained in `forecast_target_observations`, into the data frame of forecast outputs. This enables direct comparison of predictions and observations. We illustrate this here, omitting some columns from the display for convenience:

```{r}
forecast_outputs |>
  left_join(forecast_target_observations) |>
  select(-model_id, -reference_date, -horizon)
```

## The `wk flu hosp rate` target

The `wk flu hosp rate` target represents the rate of weekly confirmed influenza hospital admissions per 100,000 population. Note that this target was not included in the FluSight hub; we have introduced it here for illustrative purposes. We have used population values of 6,978,662 for Massachusetts and 29,914,599 for Texas. These population values are sourced from the [auxiliary data file](https://github.com/cdcepi/FluSight-forecast-hub/blob/main/auxiliary-data/locations.csv) provided by the FluSight hub, which are also reproduced in the [example-complex-forecast-hub](https://github.com/Infectious-Disease-Modeling-Hubs/example-complex-forecast-hub) repository.

For this target, we created cumulative distribution function (CDF) predictions with evenly spaced CDF evaluation points ranging from 0.25 to 25 in increments of 0.25 hospitalizations per 100,000 population:

```{r}
forecast_outputs |>
  filter(target == "wk flu hosp rate") |>
  select(-model_id, -reference_date, -horizon) |>
  head()
```

For the CDF `output_type`, the `output_type_id` contains the value at which the predictive CDF was evaluated, and the `value` contains the predicted probability that the target is less than or equal to that evaluation point. In the above example, the `value` in the row with `output_type_id` equal to 1.5 contains the model's estimated probability that the rate of hospital admissions in Texas the week of December 17, 2022 would be less than or equal to 1.5 admissions per 100,000 population. Again, these CDF values were estimated from the original quantile forecasts using the methods in the `distfromq` package.

The following plot illustrates the predictive CDFs produced by the `MOBS-GLEAM_FLUH` model for Massachusetts on the reference date 2022-12-17, with each `target_end_date` shown in a separate facet.  Also shown in orange is a CDF representing the observation for this target, which was between 9.75 and 10 hospitalizations per 100,000 population.  This CDF corresponds to a point mass at the observed value, with a value of 0 below the observation and a value of 1 above the observation.

```{r}
# Subset the forecasts and observations to those that we will plot
forecasts_to_plot <- forecast_outputs |>
  filter(
    model_id == "MOBS-GLEAM_FLUH",
    target == "wk flu hosp rate",
    location == "25",
    reference_date == "2022-12-17"
  ) |>
  mutate(output_type_id = as.numeric(output_type_id))
head(forecasts_to_plot)

target_observations_to_plot <- forecast_target_observations |>
  filter(
    target == "wk flu hosp rate",
    location == "25",
    target_end_date %in% forecasts_to_plot$target_end_date
  ) |>
  mutate(output_type_id = as.numeric(output_type_id))
target_observations_to_plot |>
  select(-reference_date, -horizon) |>
  head()

# We illustrate that the cdf values recorded in forecast_target_observations
# correspond to a point mass at the observed hospitalization rate.
first_one_ind <- min(which(target_observations_to_plot$observation == 1))
target_observations_to_plot[(first_one_ind - 2):(first_one_ind + 2), ]

# Make the plot
ggplot() +
  geom_line(
    mapping = aes(x = output_type_id, y = value,
                  color = "forecast", linetype = "forecast"),
    data = forecasts_to_plot
  ) +
  geom_line(
    mapping = aes(x = output_type_id, y = observation,
                  color = "observation", linetype = "observation"),
    data = target_observations_to_plot,
  ) +
  scale_color_manual(
    "CDF",
    values = c("black", "orange")
  ) +
  scale_linetype_manual(
    "CDF",
    values = c(1, 2)
  ) +
  facet_wrap(vars(target_end_date)) +
  xlab("output_type_id (units are hospital admissions per 100,000 population)") +
  ylab("CDF value (units are probability)")
```

## The `wk flu hosp rate category` target

The "wk flu hosp rate category" target represents a categorical intensity level of influenza activity, defined as "low" (hospital admissions rate per 100,000 $\leq$ 2.5), "moderate" (2.5 < admissions rate $\leq$ 5), "high" (5 < admissions rate $\leq$ 7.5), or "very high" (7.5 < admissions rate).  The `forecast_outputs` object has example forecasts for this target in a PMF format, with a probability assigned to each intensity category.  Again, forecasts of this target were not collected by the FluSight hub; we have derived predictions from the submitted quantile forecasts using the `distfromq` package.  For context, the following plot displays the observed data for the 2022/23 season on the scale of hospital admissions, with the boundaries of the intensity categories denoted with horizontal lines:

```{r}
# a data frame containing location FIPS codes and population values in units of 100,000 people
population_values <- data.frame(
  location = c("25", "48"),
  population_100k = c(6978662, 29914599) / 100000
)

# compute observed hospital admission rates for the 2022/23 season
observed_rates <- forecast_target_ts |>
  filter(location %in% c("25", "48"),
         date >= "2022-10-01", date <= "2023-04-01") |>
  left_join(population_values) |>
  mutate(rate = observation / population_100k)

# plot along with intensity thresholds
ggplot() +
  geom_line(
    mapping = aes(x = date, y = rate),
    data = observed_rates
  ) +
  geom_hline(
    mapping = aes(yintercept = y),
    linetype = 2,
    data = data.frame(y = c(2.5, 5, 7.5))
  ) +
  facet_wrap(vars(location))
```

Here is a plot showing the predictive distributions for these targets from the three included models.  Color indicates the predicted probability for each intensity category.  The observed category is indicated with a `+` in the plot, while unobserved categories are indicated with an `o`.  Here, the PMF value recorded in `forecast_target_observations` corresponds to a point mass at the observed category, with a value of 1 for the observed category and a value of 0 for the other categories.

```{r}
forecasts_to_plot <- forecast_outputs |>
  filter(
    target == "wk flu hosp rate category",
    reference_date == "2022-12-17"
  ) |>
  mutate(
    output_type_id = factor(output_type_id,
                            levels = c("low", "moderate", "high", "very high"),
                            ordered = TRUE)
  )
forecasts_to_plot |>
  select(-model_id, -reference_date, -horizon) |>
  head()

observations_to_plot <- forecast_target_observations |>
  filter(
    location %in% c("25", "48"),
    target == "wk flu hosp rate category",
    target_end_date %in% forecasts_to_plot$target_end_date
  ) |>
  mutate(
    output_type_id = factor(output_type_id,
                            levels = c("low", "moderate", "high", "very high"),
                            ordered = TRUE)
  )
observations_to_plot |>
  head()

ggplot() +
  geom_raster(
    mapping = aes(x = target_end_date, y = output_type_id, fill = value),
    data = forecasts_to_plot
  ) +
  scale_fill_viridis_c(breaks = seq(from = 0, to = 1, by = 0.2),
                       limits = c(0, 1)) +
  geom_point(
    mapping = aes(x = target_end_date, y = output_type_id, shape = factor(observation)),
    data = observations_to_plot
  ) +
  scale_shape_manual(
    values = c(1, 3),
    breaks = c(0, 1)
  ) +
  facet_grid(rows = vars(model_id), cols = vars(location)) +
  ylab("output_type_id (intensity level category)")
```
